Self-reported LLM usage and results on a data science project: Evidence from a Canadian undergraduate data science course

### 2024-08-11 paper/paper.qmd
Initial changes & added more recent chatbot references

In the context of teaching statistics and data science, LLMs could be useful for many tasks. For instance, there is considerable interest in the potential of chatbots to act as personalized tutors for students [@Winkler_2018]. ChatGPT has also been the catalyst for many interesting and important conversations around academic integrity [@Eke2023], the development of critical thinking skills, and what effective learning looks like [@BaidooAnu_2023].

@@

@Cahill_2024 surveyed undergraduate political science students on their attitudes toward and usage of AI tools. They found that the use of ChatGPT was widespread. However, they also found that many students lacked confidence in using AI for academic purposes. In particular, only 11% "strongly agreed" that they know how to use AI to improve their writing. Students had nuanced views on appropriate AI use. In particular, respondents found that using it to write whole papers as inappropriate, while using it for basic tasks like general assistance, writing feedback and basic data visualization was perceived more appropriately.

@@

All but one respondent completed the survey within 45 minutes (@fig-response-time-and-gpa-1). That one respondent took more than 4,000 minutes to complete the survey, suggesting they took a break while filling it out. That respondent is included in the analysis dataset, but without that respondent, the average time to complete the survey was 11 minutes, and the standard deviation was 8 minutes.

@@

all_data |>
  rename(ai_schoolwork_appropriate = to_what_extent_do_you_think_using_generative_ai_tools_such_as_chat_gpt_by_open_ai_or_equivalents_is_ethical_and_appropriate_for_schoolwork) |>
  mutate(
    ai_schoolwork_appropriate = case_when(
      ai_schoolwork_appropriate == "Depends of what the main goal of schoolwork is" ~ "It depends",
      ai_schoolwork_appropriate == "I believe student can only use it if they are allowed to" ~ "It depends",
      ai_schoolwork_appropriate == "a controllable amount" ~ "It depends",
      ai_schoolwork_appropriate == "appropriate depending on the context" ~ "It depends",
      ai_schoolwork_appropriate == "appropriate with rules for usage" ~ "It depends",
      ai_schoolwork_appropriate == "depends" ~ "It depends",
      ai_schoolwork_appropriate == "depends on how it is used" ~ "It depends",
      ai_schoolwork_appropriate == "half and half" ~ "It depends",
      ai_schoolwork_appropriate == "unsure" ~ "It depends",
      TRUE ~ ai_schoolwork_appropriate
    )
  ) |>
  count(ai_schoolwork_appropriate) |>
  kable(col.names = c("Ethical & Appropriate for school?", "Students"))
```

@@

all_data |>
  count(llm_usage) |> 
  kable(col.names = c("Self-reported LLM usage", "Number"))
```

@@

Define $y_i$ as the percentage received on the final paper. Then $\beta_1$ is the characterization of self-reported LLM usage and $\beta_2$ is self-reported GPA.
\begin{align}
y_i &\sim \mbox{Beta}(\mu_i, \phi) \\
logit(\mu_i) &= \beta_0 + \beta_1 \times \mbox{LLM usage}_i + \beta_2 \times \mbox{GPA}_i\\
\beta_0, \beta_1, \beta_2  &\sim \mbox{Normal}(0, 1) \\
\phi &\sim \mbox{Gamma}(4, 0.1)
\end{align}

@@

# Appendix {-}

@@

1. After carefully reading the informed consent document, please indicate below whether you consent to have your anonymized responses included in the research study?
    - Yes, I authorize the use of the data collected about me for the STA302 course survey to be used. I will be compensated 1% of my course grade for completing the survey.
    - No I do not want my data included in the research study, but I want to complete the survey. I will be compensated 1% of my course grade for completing the survey.
    - I do not want to complete this survey. I realize that I am forfeiting the corresponding course credit.
2. What is your full name on Quercus?
3. What is your Student ID?
4. What year are you?
5. What is your specialization?
6. What is/are your major/s?
7. What is/are your minor/s?
8. What is your GPA?
9. Please rate how much each statement describes you, on a scale from "This is very different to me" to "This is a lot like me" ["This is very different to me"; "This somewhat describes me"; "This is a lot like me"]
    - Writing is easy for me
    - I like to write
    - I believe it is important to be a good writer.
    - When I edit it is easy for me to catch my mistakes.
    - I feel confident sharing my writing.
    - I am confident in my overall writing ability.
10. Please rate how much each statement describes you, on a scale from "This is very different to me" to "This is a lot like me". When answering, please consider whichever programming language you are most familiar with. ["This is very different to me"; "This somewhat describes me"; "This is a lot like me"]
    - Coding is easy for me
    - I like to code
    - I believe it is important to be a good coder.
    - When I check my code it is easy for me to catch my mistakes.
    - I feel confident sharing my code.
    - I am confident in my overall coding ability.
11. How familiar are you with using generative AI tools such as OpenAI's ChatGPT or equivalents?
    - Very familiar
    - Somewhat familiar
    - Not familiar
    - Other
12. Have you used any generative AI tools such as OpenAI's ChatGPT or equivalents for any reason (personal or educational)?
    - Yes
    - No
    - Other
13. If you have used generative AI tools such as OpenAI's ChatGPT or equivalents, in what ways have you used it (select all that apply)?
    - Asking technical questions
    - Carrying on a conversation out of curiosity
    - Asking general knowledge questions
    - Solving homework
    - Checking solutions
    - Asking quick questions when stuck
    - Explaining concepts
    - Writing essays or paragraphs
    - Writing code
    - Never used it
    - Other
14. To what extent do you think using generative AI tools such as ChatGPT by OpenAI (or equivalents) is ethical and appropriate for schoolwork?
    - Appropriate
    - Inappropriate
    - Other
15. Please elaborate on your answer above.
16. Did you use any generative AI tools such as OpenAI's ChatGPT or equivalents for STA302?
    - Yes
    - No
    - Other
17. How helpful did you find generative AI tools such as ChatGPT by OpenAI (or equivalents) for each component of STA302? ["Not helpful"; "Somewhat helpful"; "Very helpful"; "I did not use generative AI for this component"]
    - Weekly quiz
    - Weekly mini-essay
    - Papers: Generating ideas
    - Papers: Writing code
    - Papers: Writing content
    - Papers: Improving content
18. What is your recommendation for how generative AI tools such as ChatGPT by OpenAI (or equivalents) should be used in the course in future?
19. (Optional) Any other comments?

>>>>

In the context of teaching statistics and data science, LLMs could be useful for many tasks. For instance, there is considerable interest in the potential of chatbots to act as personalized tutors for students [fulgencio2024developing; afzal2024tailoring; @Winkler_2018]. ChatGPT has also been the catalyst for many interesting and important conversations around academic integrity [@Eke2023], the development of critical thinking skills, and what effective learning looks like [@BaidooAnu_2023].

@@

@Cahill_2024 surveyed undergraduate political science students on their attitudes toward and usage of AI tools. They found that the use of ChatGPT was widespread. However, they also found that many students lacked confidence in using AI for academic purposes. In particular, only 11% "strongly agreed" that they know how to use AI to improve their writing. Students had nuanced views on appropriate AI use. In particular, respondents found that using it to write whole papers as inappropriate, while using it for basic tasks like general assistance, writing feedback and basic data visualization was perceived as more appropriate.

@@

All but one respondent completed the survey within 45 minutes (@fig-response-time-and-gpa-1). That one respondent took more than 4,000 minutes to complete the survey, suggesting they took a break while filling it out. That respondent is included in the analysis dataset, but without them, the average time to complete the survey was 11 minutes, and the standard deviation was 8 minutes.

@@

all_data |>
  rename(ai_schoolwork_appropriate = to_what_extent_do_you_think_using_generative_ai_tools_such_as_chat_gpt_by_open_ai_or_equivalents_is_ethical_and_appropriate_for_schoolwork) |>
  mutate(
    ai_schoolwork_appropriate = case_when(
      ai_schoolwork_appropriate == "Depends of what the main goal of schoolwork is" ~ "It depends",
      ai_schoolwork_appropriate == "I believe student can only use it if they are allowed to" ~ "It depends",
      ai_schoolwork_appropriate == "a controllable amount" ~ "It depends",
      ai_schoolwork_appropriate == "appropriate depending on the context" ~ "It depends",
      ai_schoolwork_appropriate == "appropriate with rules for usage" ~ "It depends",
      ai_schoolwork_appropriate == "depends" ~ "It depends",
      ai_schoolwork_appropriate == "depends on how it is used" ~ "It depends",
      ai_schoolwork_appropriate == "half and half" ~ "It depends",
      ai_schoolwork_appropriate == "unsure" ~ "It depends",
      TRUE ~ ai_schoolwork_appropriate
    )
  ) |>
   mutate(
     ai_schoolwork_appropriate = factor(ai_schoolwork_appropriate, levels=c("Appropriate", "It depends", "Inappropriate"))
   ) |>
  count(ai_schoolwork_appropriate) |>
  kable(col.names = c("Ethical & Appropriate for school?", "Students"))
```

@@

all_data |>
  mutate(llm_usage = factor(llm_usage, levels = c("Extensive", "Somewhat", "Minimal", "None"))) |>
  count(llm_usage) |> 
  kable(col.names = c("Self-reported LLM usage", "Number"))
```

@@

Define $y_i$ as the percentage received on the final paper. Then $\beta_1$ and $\beta_2$ are the effect of self-reported LLM usage and self-reported GPA on the final paper grade, respectively. \begin{align}
y_i &\sim \mbox{Beta}(\mu_i, \phi) \\
logit(\mu_i) &= \beta_0 + \beta_1 \times \mbox{LLM usage}_i + \beta_2 \times \mbox{GPA}_i\\
\beta_0, \beta_1, \beta_2  &\sim \mbox{Normal}(0, 1) \\
\phi &\sim \mbox{Gamma}(4, 0.1)
\end{align}

@@

# Appendix {.unnumbered}

@@

1.  After carefully reading the informed consent document, please indicate below whether you consent to have your anonymized responses included in the research study?
    -   Yes, I authorize the use of the data collected about me for the STA302 course survey to be used. I will be compensated 1% of my course grade for completing the survey.
    -   No I do not want my data included in the research study, but I want to complete the survey. I will be compensated 1% of my course grade for completing the survey.
    -   I do not want to complete this survey. I realize that I am forfeiting the corresponding course credit.
2.  What is your full name on Quercus?
3.  What is your Student ID?
4.  What year are you?
5.  What is your specialization?
6.  What is/are your major/s?
7.  What is/are your minor/s?
8.  What is your GPA?
9.  Please rate how much each statement describes you, on a scale from "This is very different to me" to "This is a lot like me" \["This is very different to me"; "This somewhat describes me"; "This is a lot like me"\]
    -   Writing is easy for me
    -   I like to write
    -   I believe it is important to be a good writer.
    -   When I edit it is easy for me to catch my mistakes.
    -   I feel confident sharing my writing.
    -   I am confident in my overall writing ability.
10. Please rate how much each statement describes you, on a scale from "This is very different to me" to "This is a lot like me". When answering, please consider whichever programming language you are most familiar with. \["This is very different to me"; "This somewhat describes me"; "This is a lot like me"\]
    -   Coding is easy for me
    -   I like to code
    -   I believe it is important to be a good coder.
    -   When I check my code it is easy for me to catch my mistakes.
    -   I feel confident sharing my code.
    -   I am confident in my overall coding ability.
11. How familiar are you with using generative AI tools such as OpenAI's ChatGPT or equivalents?
    -   Very familiar
    -   Somewhat familiar
    -   Not familiar
    -   Other
12. Have you used any generative AI tools such as OpenAI's ChatGPT or equivalents for any reason (personal or educational)?
    -   Yes
    -   No
    -   Other
13. If you have used generative AI tools such as OpenAI's ChatGPT or equivalents, in what ways have you used it (select all that apply)?
    -   Asking technical questions
    -   Carrying on a conversation out of curiosity
    -   Asking general knowledge questions
    -   Solving homework
    -   Checking solutions
    -   Asking quick questions when stuck
    -   Explaining concepts
    -   Writing essays or paragraphs
    -   Writing code
    -   Never used it
    -   Other
14. To what extent do you think using generative AI tools such as ChatGPT by OpenAI (or equivalents) is ethical and appropriate for schoolwork?
    -   Appropriate
    -   Inappropriate
    -   Other
15. Please elaborate on your answer above.
16. Did you use any generative AI tools such as OpenAI's ChatGPT or equivalents for STA302?
    -   Yes
    -   No
    -   Other
17. How helpful did you find generative AI tools such as ChatGPT by OpenAI (or equivalents) for each component of STA302? \["Not helpful"; "Somewhat helpful"; "Very helpful"; "I did not use generative AI for this component"\]
    -   Weekly quiz
    -   Weekly mini-essay
    -   Papers: Generating ideas
    -   Papers: Writing code
    -   Papers: Writing content
    -   Papers: Improving content
18. What is your recommendation for how generative AI tools such as ChatGPT by OpenAI (or equivalents) should be used in the course in future?
19. (Optional) Any other comments?
